Viscous Erosion in Porous Media

### 2018-03-04 docs/methods/erosion_methods.tex
NM

For numerical stability, we modify the erosion law \eqref{eqn:erosionModel} to include a term that penalizes curvature $\kappa$. The idea is to prevent regions of extremely high curvature from developing, as these regions would compromise the accuracy and stability of our fluid solver. We thus modify \eqref{eqn:erosionModel} follows
\begin{align}
\label{eqn:Vn}
\Vn = \atau + \eps \mean{\atau} \left( \frac{L }{2\pi} \kappa - 1 \right).
\end{align}
The parameter $\eps \ll 1$ controls the strength of the curvature-penalization. This term causes regions of high curvature to recede more rapidly and thus suppresses any high-frequency oscillations that may exist. Above, $\mean{\cdot}$ indicates a boundary mean. Since a closed body has mean curvature $\mean{\kappa} = 2\pi / L$, the penalty term has mean zero and and thus preserves area. The only source of material loss is therefore the shear-stress itself. In addition, we have scaled the penalty term on the mean absolute shear, $\mean{\atau}$, so that the ratio of smoothing-to-erosion always remains fixed, even as bodies vanish.

@@

When recast in the variables $\theta$ and $L$, the interface evolution becomes
\begin{align}
\tderiv{L}{t} &= - 2 \pi \mean{\thalpha \Vn}, \\
\pderiv{\theta}{t} &= \frac{2 \pi}{L} \left( \pderiv{\Vn}{\alpha} +
\thalpha \Vs \right).
\end{align}
Inserting~\eqref{eqn:Vn2} for the normal velocity gives
\begin{align}
\label{eqn:Levo1}
\tderiv{L}{t} &= - 2 \pi \mean{\thalpha \Vn}, \\
\label{eqn:thetaevo1}
\pderiv{\theta}{t} &= \frac{2\pi}{L} \left(
\eps \mean{\atau} \ppd{\theta}{\alpha} + \pderiv{\atau}{\alpha} +
\thalpha \Vs \right).
\end{align}
These are the main evolution equations for $\theta(\alpha,t)$ and $L(t)$.
In addition, it is necessary to track a reference point for each body, so that the {\thL} variables can be converted back to cartesian coordinates. We track the coordinates of the surface mean $\mean{\xx}$, which moves according to
\begin{align}
\label{eq:xsurf}
\frac{d}{dt} \mean{\xx} = \mean{ V_s \ss + V_n \nn }
\end{align}
We note that tracking the center of mass is also a valid choice, but the surface mean gives a somewhat simpler formula. Equations \eqref{eqn:Levo1}--\eqref{eq:xsurf} comprise the main system of equations to be solve numerically in Section~\ref{sec:timeStepping}.

@@

With the shear stress computed, we are now ready to numerically evolve the boundaries using the {\thL} framework. The main equations to be solved are \eqref{eqn:Levo1}--\eqref{eq:xsurf}.  In Section~\ref{sec:thetaL} we introduced a curvature-penalizing term into the erosion law \eqref{eqn:Vn}. This term suppresses high-frequency oscillations, but also introduces {\em stiffness} into evolution equation \eqref{eqn:thetaevo1}. We will therefore use a combined exponential-integrator/Runge-Kutta method to solve system \eqref{eqn:Levo1}--\eqref{eq:xsurf}. The exponential integrator will handle the stiff term in a numerically stable fashion, while the Runge-Kutta method will treat nonlinear terms explicitly. Our overall time-stepping method will be second order.

First, we must address a second potential source of numerical instability. The erosion law \eqref{eqn:Vn} depends on the {\em absolute value} of shear stress, $\atau$, which exhibits a corner anywhere $\tau$ changes sign (i.e.~at a stagnation point). The corner in $\atau$ introduces high frequencies into \eqref{eqn:thetaevo1} which could result in either numerical instability or an extremely stringent step-size restriction. To avoid these difficulties, we smooth $\atau$ via a narrow Gaussian filter of width $\sigma$
\begin{align}
\atausig = \int_{\gamma} \frac{1}{\sqrt{2\pi} \sigma}
 \exp \left(- \frac{\alpha'^2}{2 \sigma^2}\right) \abs{ \tau(\alpha - \alpha') } \, d\alpha' \, ,
\end{align}
with corresponding normal and tangential interface velocities
\begin{align}
& \Vnsig = \atausig +  \eps \mean{\atausig}
\left(\pderiv{\theta}{\alpha} - 1 \right), \\
& \pderiv{\Vssig}{\alpha} = \thalpha \Vnsig - \mean{\thalpha \Vnsig}.
\end{align}
This smoothing further suppresses feedback between high frequencies in $\atau$ and body geometry.

We now consider the system of evolution equations given by \eqref{eqn:Levo1}--\eqref{eq:xsurf}. For convenience, we introduce the following variables
\begin{align}
%\label{eq:M}
& \Mterm = - 2 \pi \mean{\thalpha \Vnsig}, \\
%\label{eq:N}
& \NLterm = \frac{2 \pi}{L} \left( \pderiv{\atausig}{\alpha} + \thalpha
\Vssig \right), \\
%\label{eq:elfun}
& \elfun = \frac{2 \pi}{L}  \mean{\atausig} \, .
\end{align}
Then, ~\eqref{eqn:Levo1} and~\eqref{eqn:thetaevo1} can be compactly expressed as
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo2}
& \pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + \NLterm.
\end{align}
The first term on the right-hand-side of \eqref{eqn:thetaevo2} is the stiff term resulting from the curvature penalization; i.e.~a curvature-driven interface velocity gives rise to a diffusive term for the tangent angle $\theta$. This term is multiplied by $\zeta$, which is spatially constant but depends on time (due to $L$ and $\mean{\atau}$ being time dependent). Meanwhile, $\NLterm$, represents all remaining terms in the evolution equation, including contributions from the shear-dependent erosion law and from the artificial tangential velocity, $\Vs$. Since the local shear stress depends on interactions between bodies, $\NLterm$ is both a nonlinear and nonlocal.

>>>>

For numerical stability, we modify the erosion law \eqref{eqn:erosionModel} to include a term that penalizes curvature $\kappa$. The idea is to prevent regions of extremely high curvature from developing, as these regions would compromise the accuracy and stability of our fluid solver. We thus modify \eqref{eqn:erosionModel} as follows
\begin{align}
\label{eqn:Vn}
\Vn = \atau + \eps \mean{\atau} \left( \frac{L }{2\pi} \kappa - 1 \right).
\end{align}
The parameter $\eps \ll 1$ controls the strength of the curvature-penalization. This term causes regions of high curvature to recede more rapidly and thus suppresses any high-frequency oscillations that may exist. Above, $\mean{\cdot}$ indicates a boundary mean. Since a closed body has mean curvature $\mean{\kappa} = 2\pi / L$, the penalty term has mean zero and and thus preserves area. The only source of material loss is therefore the shear-stress itself. In addition, we have scaled the penalty term on the mean absolute shear, $\mean{\atau}$, so that the ratio of smoothing-to-erosion always remains fixed, even as bodies vanish.

@@

When recast in the variables $\theta$ and $L$, the interface evolution becomes
\begin{align}
\tderiv{L}{t} &= - 2 \pi \mean{\thalpha \Vn}, \\
\pderiv{\theta}{t} &= \frac{2 \pi}{L} \left( \pderiv{\Vn}{\alpha} +
\thalpha \Vs \right).
\end{align}
Inserting~\eqref{eqn:Vn2} for the normal velocity gives
\begin{align}
\label{eqn:Levo1}
\tderiv{L}{t} &= - 2 \pi \mean{\thalpha \Vn}, \\
\label{eqn:thetaevo1}
\pderiv{\theta}{t} &= \frac{2\pi}{L} \left(
\eps \mean{\atau} \ppd{\theta}{\alpha} + \pderiv{\atau}{\alpha} +
\thalpha \Vs \right).
\end{align}
These are the main evolution equations for $\theta(\alpha,t)$ and $L(t)$.
In addition, it is necessary to track a reference point for each body, so that the {\thL} variables can be converted back to cartesian coordinates. We track the coordinates of the surface mean $\mean{\xx}$, which moves according to
\begin{align}
\label{eq:xsurf}
\frac{d}{dt} \mean{\xx} = \mean{ V_s \ss + V_n \nn }
\end{align}
We note that tracking the center of mass is also a valid choice, but the surface mean gives a somewhat simpler formula. Equations \eqref{eqn:Levo1}--\eqref{eq:xsurf} comprise the main system of equations to be solved numerically in Section~\ref{sec:timeStepping}.

@@

With the shear stress computed, we are now ready to numerically evolve the boundary of each body using the {\thL} framework. The main equations to be solved are \eqref{eqn:Levo1}--\eqref{eq:xsurf}.  In Section~\ref{sec:thetaL} we introduced a curvature-penalizing term into the erosion law \eqref{eqn:Vn}. This term suppresses high-frequency oscillations, but also introduces {\em stiffness} into evolution equation \eqref{eqn:thetaevo1}. We will therefore use a combined exponential-integrator/Runge-Kutta method to solve system \eqref{eqn:Levo1}--\eqref{eq:xsurf}. The exponential integrator will handle the stiff term in a numerically stable fashion, while the Runge-Kutta method will treat nonlinear terms explicitly. Our overall time-stepping method will be second order.

First, we must address a second potential source of numerical instability. The erosion law \eqref{eqn:Vn} depends on the {\em absolute value} of shear stress, $\atau$, which exhibits a corner anywhere $\tau$ changes sign (i.e.~at a stagnation point). The corner in $\atau$ introduces high frequencies into \eqref{eqn:thetaevo1} which could result in either numerical instability or an extremely stringent step-size restriction. To avoid these difficulties, we smooth $\atau$ via a narrow Gaussian filter of width $\sigma$
\begin{align}
\atausig = \int_{\gamma} \frac{1}{\sqrt{2\pi} \sigma}
 \exp \left(- \frac{\alpha'^2}{2 \sigma^2}\right) \abs{ \tau(\alpha - \alpha') } \, d\alpha' \, ,
\end{align}
with corresponding normal and tangential interface velocities
\begin{align}
& \Vnsig = \atausig +  \eps \mean{\atausig}
\left(\pderiv{\theta}{\alpha} - 1 \right), \\
& \pderiv{\Vssig}{\alpha} = \thalpha \Vnsig - \mean{\thalpha \Vnsig}.
\end{align}
This smoothing further suppresses the feedback of high frequencies betweeen $\atau$ and body shape.

We now consider the system of evolution equations given by \eqref{eqn:Levo1}--\eqref{eq:xsurf}. For convenience, we introduce the following variables
\begin{align}
%\label{eq:M}
& \Mterm = - 2 \pi \mean{\thalpha \Vnsig}, \\
%\label{eq:N}
& \NLterm = \frac{2 \pi}{L} \left( \pderiv{\atausig}{\alpha} + \thalpha
\Vssig \right), \\
%\label{eq:elfun}
& \elfun = \frac{2 \pi}{L}  \mean{\atausig} \, .
\end{align}
Then, ~\eqref{eqn:Levo1} and~\eqref{eqn:thetaevo1} can be compactly expressed as
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo2}
& \pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + \NLterm.
\end{align}
The first term on the right-hand-side of \eqref{eqn:thetaevo2} is the stiff term resulting from the curvature penalization; i.e.~a curvature-driven interface velocity gives rise to a diffusive term in the tangent angle $\theta$. This term is multiplied by $\zeta$, which is spatially constant but depends on time (due to $L$ and $\mean{\atau}$). Meanwhile, $\NLterm$, represents all remaining terms in the evolution equation, including contributions from the shear-dependent erosion law and from the artificial tangential velocity, $\Vs$. Since the local shear stress depends on interactions between bodies, $\NLterm$ is both a nonlinear and nonlocal.
